---
title: "Case Definition"
subtitle: "Carehome Outbreak Exercise"
---

```{r}
#| include: false
if(!require("pacman")){install.packages(pacman)}
pacman::p_load(tidyverse, rio, here, lubridate, janitor, webexercises)

clean <- import(here("data", "backup", "carehome_clean1.rds"), 
                trust = T)
```

## Set yourself up

For this exercise, you can continue working on the Data Cleaning script or create a new one solely for the task:

- **If you completed the whole cleaning process**, and are confident about your cleaning results, just continue writing a new section in your current script.

- If you could not finish it, or would like a fresh start for this task, create a second script named `02_Case_Definition`, and load the libraries you expect to need and the backup dataset provided to you in the `<backup>` data folder: `carehome_clean1.rds`

``` r
## Check wether pacman is installed and activate it or install and activate otherwise
if(!require("pacman")){install.packages(pacman)}

## Activate libraries with pacman
pacman::p_load(
  here,          # to create file paths
  rio,           # to import data
  janitor,       # tools for data cleaning and exploration
  gtsummary,     # for summary tables
  skimr,         # to explore dataframes  [NEW ONE]
  tidyverse,     # for data manipulation (always the last library!!!)
  
  update = T     # to make sure libraries are updated if new versions exist
)

## Import clean data using rio and here
clean <- import(here("data", "backup", "carehome_clean1.rds"), 
                trust = T)
```

::: {.callout-tip collapse="true"}
## Using your own cleaned data? Let's run a quick check

If you finished the cleaning exercises and want to use that data for this exercise, I invite you to load the backup data and run the following check:

``` r
# laod backup data
backup <- import(here("data", "backup", "carehome_clean1.rds"), trust = T)

# Assumen `clean` is your own dataframe from exercise 01
clean

# Are the two exactly identical?
backup == clean
```

Because we did not ask you to change any variable's name or create a new one, both dataframes should be identical if you followed the instructions correctly
:::

## 1. Agree on a Case Definition

The Case Definition is a key step in any outbreak investigation. We need to agree on the conditions that a person needs to meet in order to be considered a case, i.e., not everyone attending the Family Day is by default exposed to the food vehicle, nor all symptomatic persons are sick.

Before continuing, stop and think about a plausible Case Definition for our outbreak. It's always useful to start from the classical triad: *place, persone* and *time:*

1.  **Place**: Does the mere fact of attending the Family Day qualify you as a potential case?

2.  **Person**: Is everyone equally likely to become a case? Are all affected attendees cases by default?

3.  **Time**: Every sick person is automatically considered a case?

Take a moment to think about them. Discuss with your colleagues and facilitators around if you'd like to. But most importantly for this exercise: how would the answer to those questions translate into your existing variables? Do you have all necessary information or you need to create new ones?

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

After discussing, you finally reach the following Case Definition:

A Case is a person who:

: Attended the Family Day at "*Residencia Los Almendros*" care home on Sept 13th

  Ate from the lunch buffet

  Became sick after eating (developed any symptoms)

  Became sick within a defined time period of relevance (12-72h)

Other Case Definitions were possible, we just made a decision for this exercise. Sometimes you can develop multiple definitions to asses sensitivity and test its impact on the analyses, but that won't the be the case today
:::

## 2. Who can be a Case?

### Place criteria

Outbreaks linked to events or places may make you think the place criteria is not necessary. You ask for the attendance list and you have all potential exposed. But there is more to it when you examine the matter carefully.

**Action** — Do you have any variable that would make an attendee of the event disqualified to become a case? Think about the suspected transmission route of this outbreak

**Action** — What would you do with these persons? Can they stay in your study population?

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

The variable `buffet` identifies attendees that also ate at the lunch buffet. If someone did not eat from there, the don't have any chance of becoming a case, and therefore are not part of our study cohort and should be excluded:

``` r
# Filter away those who did not eat from the buffet
clean <- clean %>% 
  filter(buffet == "Yes")
```

**Optional**: take a quick look at those who did not eat from the buffet. In this case, they did not collect any further information through the questionnaire. But you'd be surprised about how many times they also report other information that may mislead you if you are not careful
:::

### Person criteria

Now we only have left those who attended the Family Day (i.e., are on the linelist) and ate from the buffet. Now you need to define who actually became sick. Since we don't have any column that give you the information directly, we need to infer from the available symptoms.

**Action** — Create a new variable, `gastro_symptoms`, that becomes `1` for anyone reporting at least one of the five gastrointestinal symptoms, and `0` otherwise. Use `case_when()` for this task.

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

``` r
# Filter away those who did not eat from the buffet
clean <- clean %>%
  mutate(gastro_symptoms = case_when(
     # Use OR criteria to capture any of the symptoms as possitive condition
     diarrhoea == "Yes" | vomiting == "Yes" | 
       abdo == "Yes" | fever == "Yes" | 
       nausea == "Yes"    ~ 1,
     
     # If none of the above conditions are met, then
     .default = 0
    )
  )
```

**Stop and think**: Was including all gastro symptoms into our case definition the right choice? One should think that feeling a bit of abdominal pain, nausea, or even vomiting after eating in a buffet could mean only that you ate too much, for example. Those are very unspecific symptoms in terms of identifying potential cases given that we suspect something else going on. Once again, this is only an exercise, but always think about the consequences of your choices
:::

### Time criteria

If our hypothesis is that some consumed food was contaminated and caused an intoxication, then clinical symptoms should occur after the food was consumed. Not only that, disease should occur within a plausible time window that will later help us narrow down the potential causative agents, given their known incubation periods.

#### Define the onset day and time of sick people

For now, we have data on the day and hour of occurrence, but over two separated columns we need to merge together:

**Action** — Build a single `onset_datetime` from `dayonset` and `starthour`. For that, we have two potential options:

- `dmy_h()`, a `lubridate` function where the "h" stands for "hour", turns a character string into a date-time format. Useful if first you want to glue together both columns or modify the characters beforehand.

- If both variables are already in `<date>` format, you can just add them together with `+` and voilà.

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

``` r
# Create a date-time variable
clean <- clean %>% 
  mutate(onset_datetime = dayonset + starthour)
```
:::

**Which path should you choose**? For that, you would need to first check whether every onset day is also accompanied by an onset time. When you add them together, missing any of them will result in `NA`. What to do if some information is missing is also part of your decisions as epidemiologist. *Do you want to assume some onset time, i.e. 00h*? *Do you exclude them from the time criteria*? *Do you restrict the time criteria to whole days*? Think about it

#### Calculate incubation period in hours

The incubation period is the time elapsed between exposure to the causative agent and disease onset. It can be calculated in days or hours, depending on the pathogen and disease presentation. Since people became sick right after the buffet, hours seem like the logical choice:

**Action** — The buffet was served on **September 13th at 14:00**. Calculate the `incubation` variable, i.e. how many *hours* passed between the buffet and their symptom onset. For that:

- Create a new date-time variable `buffet_datetime` using the adequate `lubridate` function.

- Calculating the time period between any two dates can be as simple as calculating the difference between them using `-`, the subtraction symbol.

- Convert this substraction to `<numeric>` class, as `<difftime>` is a specific variable type that can be problematic to handle.

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

``` r
# Calculate the time span between two dates (Diffdate)
clean <- clean %>% 
  mutate(buffet_datetime = ymd_h("2026-09-13 14h"),
         incubation = as.numeric(onset_datetime - buffet_datetime))
```
:::

**Optional** — Explore the new `incubation` period variable. Can you find any strange values? We should now plot the incubation periods with a histogram to get an idea about our time criteria for deciding who is a case and who isn't, *developing symptoms between 12-72h post-buffet*.

```{r}
#| include: false
clean <- clean %>%
  mutate(gastro_symptoms = case_when(
    # Use OR criteria to capture any of the symptoms as possitive condition
    diarrhoea == "Yes" | vomiting == "Yes" | 
      abdo == "Yes" | fever == "Yes" | 
      nausea == "Yes"    ~ 1,
    
    # If none of the above conditions are met, then
    .default = 0
  )) %>% 
  mutate(onset_datetime = dayonset + starthour)  %>% 
  mutate(buffet_datetime = ymd_h("2026-09-13 14h"),
         incubation = as.numeric(onset_datetime - buffet_datetime))
```

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

**Interpretation**: among those developing symptoms, most of the report disease onset between 12 and 42 hours after eating the lunch buffet. Some people reported symptoms starting **before** the buffet, and also quite late \>72h post-buffet.

```{r}
clean %>% 
  ggplot(aes(x = incubation)) +
  geom_histogram(binwidth = 6) +
  scale_x_continuous(breaks = seq(min(clean$incubation, na.rm = T), 
                                  max(clean$incubation, na.rm = T), 
                                  12)
                     ) +
  labs(title = "Distribution of Incubation Periods") +
  theme_bw()
```
:::

### 3. Create the `Case` variable

Following the Case Definition we agreed upon, and after creating all needed variables, you are ready to create a new `Case` variable taking the value `1` for those meeting the case definition, and `0` otherwise

**Action** — Use `case_when()` to define your `case` variable using the person and time criteria

- How many cases fulfill the Case Definition? `r fitb(49)`
- What is the overall attack rate of this outbreak? `r fitb(29.35, num = T, tol = 0.05)`

::: {.callout-caution collapse="true" appearance="minimal" icon="false"}
## See the solution

```{r}
clean <- clean %>% 
  mutate(case = case_when(
    # They experienced gastrointestinal symptoms
    gastro_symptoms == 1 & 
      
    # Symptom onset between 12 and 72h after eating at the buffet   
    incubation >= 12 & incubation <= 72   ~ 1,
    
    # Everything else
    .default = 0
  ))

# Count and Percentage of cases vs no cases
tabyl(clean$case)
```
:::

## Save and Export the final dataframe

After so much work it's worth saving our dataframe once again. This is the "final" cleaned dataframe we can run all kind of exploration, description and anlayses in the following exercises

**Action** — Save the cleaned dataframe as `carehome_clean2.rds` in the clean data folder

``` {.r code-fold="T"}
export(clean, here("data", "clean", "carehome_clean2.rds"))
```
